---
title: "Palmer-Archipelago-Penguin Data wrangling"
author: "Bryan Brugal"
date: "4/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load the library for tidyverse package
```{r}
library(tidyverse)
```

### Load the csv file
```{r}
## Storing the data-set into a name
penguins_raw <- read.csv("penguins_raw.csv")
head(penguins_raw)
```

View(penguins_raw)

```{r}
# check for variable name 
colnames(penguins_raw)
```

I will target variables: 
- species  
- island  
- Culmen.Length..mm.   
- Culmen.Depth..mm.  
- Flipper.Length..mm. 
- Body.Mass..g  
- Sex
I should subset a new dataframe from the original dataframe that contains ONLY the variable that I targeted.  
Based on the location of the column, then I know that I need column numbers: 3, 5, 10 to 14
Subset a new dataset  
        original-data-set-name [ row  ,  column ]
        
```{r}
penguins <- penguins_raw[  , c(3, 5, 10 : 14)]
head(penguins)
tail(penguins)
```
Check for the structure

```{r}
penguins %>% str()
```

Changing variable names  
    rename ( new-name = old-name )
    
```{r}
penguins <- penguins %>% rename( 
  species = Species ,
  island = Island ,
  bill.len.mm = Culmen.Length..mm. ,
  bill.depth.mm = Culmen.Depth..mm. , 
  flipper.mm = Flipper.Length..mm. ,
  body.g = Body.Mass..g. ,
  sex = Sex ) 
penguins %>% str()
```

Check for unique values
```{r}
unique(penguins$species)
unique(penguins$island)
unique(penguins$sex)
```

Factor qualitative values
```{r}
penguins$species <- factor(penguins$species)
penguins$island <- factor(penguins$island)
penguins$sex <- factor(penguins$sex)
penguins %>% str()
```

Extract the first words of the species value to replace the original value
      first word extraction from a string value:  str_extract ( data $ variable, "\\w+"  )   
      last word extraction from a string value:  str_extract ( data $ variable, "\\w+$"  )
      
```{r}
penguins$species <- str_extract(penguins$species, "\\w+"  )
unique(penguins$species)
penguins %>% str()
penguins$species <- factor(penguins$species)
levels(penguins$species)
```

## count() - count rows/observations
Number of penguins per species
```{r}
penguins %>% count(species)
```


Number of penguins per species, island, sex 
```{r}
penguins %>% count(species, island, sex)
```

Filtering penguins from species "Chinstrap"
```{r}
penguins %>% filter( species == "Chinstrap" )
```

Penguins from EITHER Chinstrap or Gentoo
```{r}
penguins %>% filter(species == "Chinstrap" | species == "Gentoo")
```
Penguins with species Adelie AND island Dream
```{r}
penguins %>% filter(species == "Adelie" & island == "Dream")
```
```{r}
#Penguins with species Adelie AND island Biscoe
penguins %>%
  filter(species == "Adelie" & island == "Biscoe")
```

Penguin with body mass is greater than 3200g
```{r}
penguins %>% filter(body.g > 3200)
```
 Penguins where species is NOT Chinstrap
 
```{r}
penguins %>% filter(species != "Chinstrap")
```
 
  
**select()** 
Select column species and sex
```{r}
penguins %>% select(species, sex)
```
  View(penguins)  
Select species, and from bill length to sex 
```{r}
penguins %>% select(species, bill.len.mm : sex)
```
  
 Without the island column
 
```{r}
# method 1
penguins %>% select(!c(island))

```
 
```{r}
# method 2
penguins %>% select(-island)
```
 
 Columns start with "bill"
 
```{r}
penguins %>% select(starts_with("bill"))
```
 
 Columns end with "mm"
 
```{r}
penguins %>% select(ends_with("mm"))
```
 
Column that contain "len" 
```{r}
penguins %>% select(contains("len"))
```

**select()** and **filter()** 
- select and filter a subset that penguin species is gentoo, and keep the body mass and sex variable
```{r}
# check structure of the current target dataset
str(penguins)
unique(penguins$species)
# filter species = gentoo, select body mass and sex
penguins %>% 
  filter(species == "Gentoo") %>% 
  select(body.g , sex)
```

**relocate()** - move columns around, without messing with rows and groups.   
BY DEFAUT, relocate() will move an indicated column to the very first column
```{r}
head(penguins)
 # move the sex column before the species 
penguins %>% 
  relocate(sex)
```

 Move the flipper.mm before the bill.len.mm using helper `.before=`
 
```{r}
penguins %>% 
  relocate(flipper.mm, .before= bill.len.mm)
```
 
move bill.len.mm after the sex column using `.after=` 
```{r}
penguins %>% 
  relocate(bill.len.mm , .after= sex)
```

 Move the flipper.mm before the bill.len.mm using helper `.before=`
 
```{r}
penguins %>% 
  relocate(flipper.mm, .before= bill.len.mm)
```
 
move bill.len.mm after the sex column using `.after=` 
```{r}
penguins %>% 
  relocate(bill.len.mm , .after= sex)
```

## 12. mutate() -  add a new column while keeping the existing columns.   
      mutate ( new-column-name = what-it-contains ) 
      
      mutate( D = B + C )  
      
      
Add a new column which penguins body mass but in kilograms.  
g / 1000 = kg 
```{r}
penguins %>% 
  mutate(body.kg = body.g / 1000)
```
Add a new column as ratio of bill.len.mm to bill.depth.mm  
ratio = bill.len.mm / bill.depth.mm  
```{r}
penguins %>% 
  mutate(bill.ratio = bill.len.mm/bill.depth.mm)
```

```{r}
penguins %>% 
  mutate(bill.ratio = bill.len.mm/bill.depth.mm ,
         body.kg = body.g / 1000 , 
         flipper.m = flipper.mm / 1000)
```

# 13. group_by() and summarize() - find the summary statistics for different groups and put them into a nice table   
      group_by()  -  group the data   
      summarize() -  calculate something   
Generate a summary table containing the mean of penguin body mass, grouped by species
      
```{r}
penguins %>% 
  group_by(species) %>% 
  summarize(mass_mean = mean(body.g , na.rm = T))
```
      
Generate a summary table containing the mean and standard deviation of penguin body mass, grouped by species
```{r}
penguins %>% 
  group_by(species) %>% 
  summarize(mass_mean = mean(body.g , na.rm = T), 
            mass_sd = sd(body.g ,na.rm = T ))
```
Based on the result, the body mass of different penguin species are:  
- Adelie (M = 3700.662, SD = 458.5661)   
- Chinstrap (M = 3733.088, SD = 384.3351)  
- Gentoo (M = 5076.016, SD = 504.1162)


Generate a summary table containing the maximum and minimum flipper.mm for male Adelie penguins, group by the island.  
```{r}
unique(penguins$sex)
str(penguins)
penguins %>% 
  filter(species == "Adelie" , sex == "MALE") %>% 
  group_by(island) %>% 
  summarize(flipper.max = max(flipper.mm),
            flipper.min = min(flipper.mm))
```

## 13. Conditioning with `case_when()` - used within `mutate()`  
 mutate( new-variable-name = case_when( condition1 ~ value1 , condition2 ~ value2 , .... ))
 
 new column size based on body.g   
 
 - body.g > 4500 --> "large"  
 - 3000 > body.g <= 4500 --> "medium"  
 - body.g <= 3000 --> small 
 
```{r}
penguins %>% 
  mutate(size = case_when( body.g > 4500 ~ "large" ,
                           body.g > 3000 & body.g <= 4500 ~ "medium" ,
                           body.g <= 3000 ~ "small"))
```
 
# Part 4: Invalid Input and Missing Value
```{r}
unique(penguins$sex)
#conver factor sex into character
penguins$sex <- as.character(penguins$sex)
# filter the rows that contain only FEMALE and MALE
penguins <- penguins %>%
  filter(sex == "FEMALE" | sex == "MALE")
# factor variable sex again
penguins$sex <- factor(penguins$sex)
# structure of penguins
str(penguins)
```
**na.omit()** - to drop missing values. 
```{r}
unique(is.na(penguins))
penguins <- penguins %>%
  na.omit()
```


# Part 5: Visualization 
      ggplot( data = dataset-name , mapping = ase( x =  , y= )  )  +
          geom_bar() / geom_point() /geom_histogram() /geom_boxplot() + 
          labs( x = "Vertical Axis title ", 
                y = " Horizontal  Axis tile" , 
                title = "Plot Title" )
Bar Plot of Penguins Counts for different Sex 
```{r}
colors()
```

```{r}
ggplot(data = penguins , 
       mapping = aes(x = sex)) +
  geom_bar(color = "darkblue", 
           fill = "skyblue", 
           width = 0.5) +
  geom_text(stat ="count", 
            mapping = aes(label = ..count..), 
            vjust = 5) + 
  labs(x = "Penguins Sex" , 
       y = "Penguins Counts" , 
       title = "Bar plot of Penguins Counts of Different Sex") +
  theme_classic()
```

```{r}
ggplot(data = penguins, 
       mapping = aes(x = island)) +
  geom_bar(color = "purple", 
           fill = "lavender", 
           width = 0.3) +
  geom_text(stat ="count", 
            mapping = aes(label = ..count..), 
            vjust = 5) + 
  labs(x = "Island" , 
       y = "Penguins Counts" , 
       title = "Bar Graph of Penguins in Different Island") +
  theme_classic()
```

